---
title: "Compare, Test, and Select Models"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, performance, r2]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Compare, Test and Select Models}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r , include=FALSE}
library(knitr)
library(performance)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(
  comment = ">",
  message = FALSE,
  warning = FALSE,
  out.width = "100%",
  dpi = 450
)
options(digits = 2)

set.seed(333)

library(rstanarm)
```

# Comparing vs. Testing

Let's imagine that we are interested in explaining the variability in the `Sepal.Length` using 3 different predictors. For that, we can build 3 linear models.

```{r}
model1 <- lm(Sepal.Length ~ Petal.Length, data = iris)
model2 <- lm(Sepal.Length ~ Petal.Width, data = iris)
model3 <- lm(Sepal.Length ~ Sepal.Width, data = iris)
```

## Comparing Indices of Model Performance

The eponymous function from the package,
[`performance()`](https://easystats.github.io/performance/reference/model_performance.html),
can be used to compute different indices of performance (an umbrella term for
indices of fit).

```{r}
library(performance)
library(insight)
library(magrittr) # for pipe operator

# we will use `print_md` function to display a well-formatted table
performance(model1) %>%
  print_md()
```

But for multiple models, one can obtain a useful table to compare these indices
at a glance using the
[`compare_performance()`](https://easystats.github.io/performance/reference/compare_performance.html)
function.

```{r}
compare_performance(model1, model2, model3) %>%
  print_md()
```

If you remember your stats lessons, while comparing different model fits, you
would like to choose a model that has a high $R^2$ value (a measure of how much
variance is explained by predictors), low AIC and BIC values, and low root mean
squared error (RMSE). Based on these criteria, we can immediately see that
`model1` has the best fit.

If you don't like looking at tables, you can also plot them using a plotting method supported in `see` package:

```{r}
library(see)

plot(compare_performance(model1, model2, model3))
```

For more, see: <https://easystats.github.io/see/articles/performance.html>

## Testing Models

While **comparing** these indices is often useful, making a decision (for
instance, which model to keep or drop) can often be hard, as the indices can
give conflicting suggestions. Additionally, it is sometimes unclear which index
to favour in the given context.

This is one of the reason why **tests** are useful, as they facilitate decisions
via (infamous) "significance" indices, like *p*-values (in frequentist
framework) or [Bayes Factors](https://easystats.github.io/bayestestR/articles/bayes_factors.html) (in
Bayesian framework).

```{r}
test_performance(model1, model2, model3) %>%
  print_md()
```

However, these tests also have strong limitations and shortcomings, and cannot
be used as the **one criterion to rule them all**!

You can find more information on how these tests [**here**](https://easystats.github.io/performance/reference/test_performance.html).

## Experimenting

Although we have shown here examples only with simple linear models, we will
highly encourage you to try these functions out with models of your choosing.
For example, these functions work with mixed-effects regression models, Bayesian
regression models, etc.

To demonstrate this, we will run Bayesian versions of linear regression models
we just compared:

```{r}
library(rstanarm)

model1 <- stan_glm(Sepal.Length ~ Petal.Length, data = iris, refresh = 0)
model2 <- stan_glm(Sepal.Length ~ Petal.Width, data = iris, refresh = 0)
model3 <- stan_glm(Sepal.Length ~ Sepal.Width, data = iris, refresh = 0)

compare_performance(model1, model2, model3) %>%
  print_md()
```

Note that, since these are Bayesian regression models, the function
automatically picked up the appropriate indices to
compare!

If you are unfamiliar with some of these, explore more [here](https://easystats.github.io/performance/reference/looic.html).

**Now it's your turn to play!** :)
