---
title: "Checking simulated residuals"
output: 
  rmarkdown::html_vignette:
    toc: true
tags: [r, performance]
vignette: >
  \usepackage[utf8]{inputenc}
  %\VignetteIndexEntry{Checking simulated residuals}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r , include=FALSE}
library(knitr)
knitr::opts_chunk$set(
  dpi = 300,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
options(knitr.kable.NA = "")
options(digits = 2)

pkgs <- c("DHARMa", "glmmTMB", "see")
successfully_loaded <- vapply(pkgs, requireNamespace, FUN.VALUE = logical(1L), quietly = TRUE)
can_evaluate <- all(successfully_loaded)

if (can_evaluate) {
  knitr::opts_chunk$set(eval = TRUE)
  vapply(pkgs, require, FUN.VALUE = logical(1L), quietly = TRUE, character.only = TRUE)
} else {
  knitr::opts_chunk$set(eval = FALSE)
}
```

The basic workflow for simulated residual checks using `simulate_residuals()` is as follows.

First, fit a model:

```{r}
model <- glmmTMB::glmmTMB(
  count ~ mined + spp + (1 | site),
  family = poisson,
  data = glmmTMB::Salamanders
)
```

Next, simulate residuals from the model:

```{r}
library(performance)
simulated_residuals <- simulate_residuals(model)

simulated_residuals
```

The raw residuals can be extracted using `residuals()`:

```{r}
head(residuals(simulated_residuals))
```


<!-- Aside -->
Note that since this inherits the DHARMa class, all the methods implemented in DHARMa just work, including all the tests:

```{r}
DHARMa::testUniformity(simulated_residuals, plot = FALSE)
```
<!-- Aside -->

Finally, run specific checks on the simulated residuals:

```{r message=TRUE}
check_residuals(simulated_residuals)
```

Further implemented checks are tests for overdispersion, outliers and zero-inflation.

```{r message=TRUE}
check_overdispersion(simulated_residuals)

check_zeroinflation(simulated_residuals)

check_outliers(simulated_residuals)
```

The above three functions internally call `simulate_residuals()` for more complex models automatically, so you don't need to call `simulate_residuals()` yourself. Simulated residuals are usually more reliable than the standard residuals, especially for complex models.

Finally, you can even perform a visual check for the entire model, either by passing the model object directly, or the object returned from `simulate_residuals()`. 

```{r fig.height=12, fig.width=10}
check_model(simulated_residuals)
```

The `check_model()` function is the main reason we don't want to prematurely extract the residuals in `simulate_residuals()`, because if we do then the simulated residual won't contain the model fit (`fittedModel` in the output below), so we won't be able to do all of the checks we would want to do using the model (e.g., posterior predictive checks).

```{r}
str(simulated_residuals, max.level = 1)
```
